---
title: "LLMs for Quantitative Analysis — Worked Solutions"
author: "Dr Alex Goddard"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 8, fig.height = 4.8)
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(effectsize)

PAL <- c(UK = "#4C72B0", EU = "#DD8452", International = "#55A868")
theme_set(theme_minimal(base_size = 12))
```

> **The core principle: fail fast.** It is better to *continually inspect* the
> data and your results than to try to write one perfect prompt. Work
> step-by-step, run the code, look at what happens, and repeat. An LLM will
> happily give you a confident, wrong answer — your job is to catch it.

This notebook is the worked solution to the two data tasks: **cleaning** a raw
Qualtrics export, then **analysing** it. All code is R. Each step shows the
*problem*, how to *detect* it, and the *fix*.

---

# Part 1 — Data Cleaning

A survey on the wellbeing of university students, exported straight from
Qualtrics. Our goal: get from the raw export to a dataset that is ready for
analysis. There are **10 problems** — five with *rows*, five with *data entry*.

```{r load}
# Read EVERYTHING as text first — the embedded label row would otherwise break
# the column types.
raw <- read_csv("data/cultural_wellbeing_study.csv",
                col_types = cols(.default = col_character()))

dim(raw)   # 306 rows (incl. a label row) x 36 columns
```

A first look shows the trouble: row 1 is question text, gender is encoded every
which way, age contains "years", and the scale items hold odd values.

```{r peek}
raw[1:3, c("participant_id", "Status", "Finished", "gender", "age", "w1", "w4")]
```

```{r scale-cols}
# The 16 survey items: 7 wellbeing (1-5), 5 belonging (1-7), 4 discrimination (1-5)
W <- paste0("w", 1:7)
B <- paste0("b", 1:5)
D <- paste0("d", 1:4)
scale_cols <- c(W, B, D)

df <- raw   # we will whittle `df` down step by step
```

## Row problems (5)

We track the **row count** after each fix. Raw = 306; target = 299.

### Problem 1 — Row 0 is question text

*Detection:* visual inspection — the first data row contains the survey
questions, not a participant. *Fix:* drop it.

```{r row1}
df <- df %>% filter(participant_id != "Participant ID")
nrow(df)   # 305
```

### Problem 2 — Three participants did not finish

*Detection:* the `Status` / `Finished` / `Duration` columns. One row is a
researcher **preview** (a 28-second "response"), and two participants quit
part-way (`Finished == "False"`). *Fix:* keep only completed, real responses.

```{r row2}
table(df$Status, df$Finished)

df <- df %>% filter(Status == "IP Address", Finished == "True")
nrow(df)   # 302
```

### Problem 3 — Attention-check failure

The survey asked respondents to "select the number four". *Detection:* anyone
not answering 4. *Fix:* drop them.

```{r row3}
df <- df %>% mutate(attention_check = as.numeric(attention_check))
table(df$attention_check)

df <- df %>% filter(attention_check == 4)
nrow(df)   # 301
```

### Problem 4 — A straight-line responder (P301)

One participant gave the **same answer to all 16 items**. They pass every other
check — the only way to catch them is to compute the **standard deviation across
each row**: a row SD of 0 means no variation at all.

```{r row4}
# Items must be numeric to compute an SD ("N/A" strings become NA here)
df <- df %>% mutate(across(all_of(scale_cols), as.numeric))

row_sd <- apply(df[scale_cols], 1, sd, na.rm = TRUE)
df$participant_id[which(row_sd == 0)]            # P301

df <- df[!(!is.na(row_sd) & row_sd == 0), ]
nrow(df)   # 300
```

### Problem 5 — A duplicate participant (P047)

*Detection:* a repeated `participant_id`. *Fix:* keep the first occurrence.

```{r row5}
df$participant_id[duplicated(df$participant_id)]  # P047

df <- df %>% distinct(participant_id, .keep_all = TRUE)
nrow(df)   # 299  <- final N
```

## Data-entry problems (5)

These are the same five issues from the slide. We clear the **missing-value
codes and out-of-range values first**, so that the reliability check in the last
step is meaningful.

### Problem 6 — Missing data coded three different ways

Missingness hides behind `-99`, the string `"N/A"`, and true blanks. *Fix:*
unify them all to `NA`. (The `"N/A"` strings already became `NA` when we
converted the items to numeric above; here we mop up the `-99` sentinels.)

```{r data6}
sum(df[scale_cols] == -99, na.rm = TRUE)         # how many -99 sentinels?

df <- df %>% mutate(across(all_of(scale_cols), ~ na_if(.x, -99)))
sum(is.na(df[scale_cols]))                        # total missing now unified
```

### Problem 7 — Out-of-range values

A `6` appears on the 1–5 wellbeing items, a `0` on the 1–5 discrimination
items. *Detection:* inspect each item's range. *Fix:* set impossible values to
`NA`.

```{r data7}
scale_max <- c(setNames(rep(5, 7), W), setNames(rep(7, 5), B), setNames(rep(5, 4), D))
for (c in scale_cols) {
  df[[c]][df[[c]] < 1 | df[[c]] > scale_max[[c]]] <- NA
}
# confirm every item is now within range
sapply(scale_cols, function(c) range(df[[c]], na.rm = TRUE))[, 1:6]
```

### Problem 8 — Inconsistent gender encoding

*Detection:* visual inspection — `F`, `f`, `female`, `FEMALE`, `M`, `m`, `NB`…
*Fix:* a standardised mapping.

```{r data8}
table(df$gender)

df <- df %>%
  mutate(gender = str_to_lower(str_trim(gender)),
         gender = recode(gender,
                         "f" = "female",      "female" = "female",
                         "m" = "male",        "male"   = "male",
                         "nb" = "non-binary", "non binary" = "non-binary",
                         "non-binary" = "non-binary"),
         gender = str_to_title(gender))
table(df$gender)
```

### Problem 9 — Age stored as text ("24 years")

*Fix:* strip the word "years" and convert to a number.

```{r data9}
df <- df %>% mutate(age = as.numeric(str_remove(age, "\\s*years?")))
range(df$age)
```

### Problem 10 — Reverse-coded items (w4 and w6)

Two wellbeing items are **negatively worded** ("struggling to deal with
problems", "feeling distant"), so a high score means *low* wellbeing. Averaged
as-is, the scale is incoherent. *Detection:* Cronbach's α is very low; those
items correlate negatively with the rest. *Fix:* reverse-score them (`6 - old`).

```{r data10}
cronbach <- function(mat) {
  mat <- mat[complete.cases(mat), , drop = FALSE]; k <- ncol(mat)
  round((k / (k - 1)) * (1 - sum(apply(mat, 2, var)) / var(rowSums(mat))), 2)
}

cronbach(as.matrix(df[W]))               # BEFORE: ~0.21  (unacceptable)

df <- df %>% mutate(w4 = 6 - w4, w6 = 6 - w6)

cronbach(as.matrix(df[W]))               # AFTER:  ~0.89  (excellent)
```

## Build aggregate scores and save

With the items clean and reverse-coded, average them into one score per scale.

```{r composites}
df <- df %>%
  mutate(year_of_study = as.numeric(year_of_study),
         wellbeing      = rowMeans(across(all_of(W)), na.rm = TRUE),
         belonging      = rowMeans(across(all_of(B)), na.rm = TRUE),
         discrimination = rowMeans(across(all_of(D)), na.rm = TRUE))

# reliability of all three scales
c(wellbeing      = cronbach(as.matrix(df[W])),
  belonging      = cronbach(as.matrix(df[B])),
  discrimination = cronbach(as.matrix(df[D])))
```

```{r save, eval=FALSE}
# Save the cleaned dataset (already provided in /data for the analysis task)
write_csv(df, "data/cultural_wellbeing_study_clean.csv")
```

**Done — N = 299, clean and analysis-ready.** Note how much harder the cleaning
was than the analysis that follows.

---

# Part 2 — Data Analysis

> For this task the data already **contains aggregate scores** and you may
> **assume data integrity is good**. Three questions, worked one-by-one.

```{r prep-analysis}
df$group  <- factor(df$group,  levels = c("UK", "EU", "International"))
df$gender <- factor(df$gender, levels = c("Female", "Male", "Non-Binary"))

# Descriptives by group
df %>% group_by(group) %>%
  summarise(across(c(wellbeing, belonging, discrimination),
                   list(mean = ~round(mean(.x), 2), sd = ~round(sd(.x), 2))),
            n = n())
```

## Q1 — Do student groups differ in wellbeing?

A one-way **ANOVA** asks whether the three group means differ; **Tukey HSD**
tells us *which* pairs differ.

```{r q1-anova}
aov_wb <- aov(wellbeing ~ group, data = df)
summary(aov_wb)
eta_squared(aov_wb)
TukeyHSD(aov_wb)
```

```{r q1-d}
# Cohen's d for each pair
cohens_d <- function(a, b) {
  x <- df$wellbeing[df$group == a]; y <- df$wellbeing[df$group == b]
  round((mean(x) - mean(y)) / sqrt((var(x) + var(y)) / 2), 2)
}
c(`UK vs Intl` = cohens_d("UK", "International"),
  `UK vs EU`   = cohens_d("UK", "EU"),
  `EU vs Intl` = cohens_d("EU", "International"))
```

```{r q1-plot, fig.height=4.5}
ggplot(df, aes(group, wellbeing, fill = group)) +
  geom_violin(alpha = .55, colour = NA) +
  geom_boxplot(width = .15, outlier.size = .6) +
  scale_fill_manual(values = PAL, guide = "none") +
  labs(title = "Wellbeing by student group", x = NULL, y = "Wellbeing (1-5)")
```

**Finding.** F(2, 296) = 13.5, *p* < .001, η² = 0.08 (medium). Means: UK 3.59 >
EU 3.28 > International 3.12. UK differs from both EU (*p* = .003, *d* = 0.46)
and International (*p* < .001, *d* = 0.70) — **but EU vs International is *not*
significant** (*p* = .28). A significant ANOVA does not mean every pair differs.

## Q2 — Is perceived discrimination related to wellbeing?

```{r q2-cor}
round(cor(df[c("wellbeing", "belonging", "discrimination")]), 2)

summary(lm(wellbeing ~ discrimination, data = df))
```

```{r q2-plot, fig.height=4.8}
ggplot(df, aes(discrimination, wellbeing)) +
  geom_point(aes(colour = group), alpha = .5) +
  geom_smooth(method = "lm", colour = "black", linewidth = .8) +
  scale_colour_manual(values = PAL, name = NULL) +
  labs(title = "Wellbeing vs perceived discrimination",
       x = "Perceived discrimination (mean)", y = "Wellbeing (mean)")
```

**Finding.** r(wellbeing, discrimination) = −0.41; r(wellbeing, belonging) =
+0.51; r(belonging, discrimination) = −0.52 (all *p* < .001). In a simple
regression, each one-point rise in discrimination predicts a **0.34-point drop**
in wellbeing.

## Q3 — What are the key predictors of wellbeing?

Put everything in one **multiple regression**: the two other scales, group, and
the demographics (age, gender, year).

```{r q3-model}
m <- lm(wellbeing ~ belonging + discrimination + group + age + gender + year_of_study,
        data = df)
summary(m)
```

**Finding.** Only **belonging** (*B* = +0.24, *p* < .001) and **discrimination**
(*B* = −0.16, *p* = .002) predict wellbeing. Once they are in the model, **group
is no longer significant**, and none of the demographics matter. Adjusted
R² = 0.29. The group differences in Q1 are largely *explained* by belonging and
discrimination — group membership was a proxy for the underlying experience.

---

# Conclusion

- **Fail fast.** Continually inspect the data and results rather than perfecting
  a single prompt.
- **Work step-by-step** so you understand what each fix is doing.
- An LLM can also help **check the assumptions** of these tests (not covered
  here).
- **Data cleaning is harder than the analysis** — and it is where most errors,
  and most of an LLM's confident mistakes, are born.
